### 2025-08-26 sections/2-methods.tex
tightening discussion

Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can simply say that the dataset has low LFD.
We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset that contains genomic sequences with unaligned lengths of up to 3,712 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,712 and at most 50,000; as used in this paper it is 3,712.
However, physical constraints (namely, biological evolution and biochemistry) constrain the data to a lower-dimensional manifold within this space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
Section~\ref{sec:results:lfd-of-datasets} discusses this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.
For real-world datasets, we expect the LFD to be locally uniform, i.e.,\,when $r$ is small, but potentially highly variable at larger length scales, i.e.,\,when $r$ is large.

@@

We start by performing a divisive hierarchical clustering on the dataset using CLAM to obtain a cluster tree $\mathcal{T}$.
The procedure is almost identical to that outlined in CHESS~\cite{ishaq2019clustered}, but with better selection of poles for partitioning (see Algorithm~\ref{alg:methods:partition}).
After building $\mathcal{T}$ with CLAM, we also perform a depth-first reordering of the dataset (see Section~\ref{sec:methods:clustering:depth-first-reordering}).

@@

Given a cluster $C$ with $|C|$ points, we define its two children by the following process. We take a random subsample $S$ of $\sqrt{|C|}$ of $C$'s points, and compute pairwise distances between all points in $S$.
Using these distances, we compute the \textit{geometric median} of $S$; in other words, we find the point that minimizes the sum of distances to all other points in $S$.
We define the \textit{center} of $C$ to be this geometric median.

The \textit{radius} of $C$ is the maximum distance from the center to any other point in $C$.
The point that is responsible for that radius (i.e.,\,the furthest point from the center) is designated the \textit{left pole} and the point that is furthest from the left pole is designated the \textit{right pole}.
We then partition the cluster into a \textit{left child} and a \textit{right child}, where the left child contains all points in the cluster that are closer to the left pole than to the right pole, and the right child contains all points in the cluster that are closer to the right pole than to the left pole.
Without loss of generality, we assign to the left child those points that are equidistant from the two poles.
Starting from a root cluster $\mathcal{R}$ containing the entire dataset, we repeat this procedure until each leaf contains only one datum, or we meet some other user-specified stopping criteria (e.g., minimum cluster radius, minimum cluster cardinality, maximum tree depth, etc).
This process is described in Algorithm \ref{alg:methods:partition}.
During the partitioning process, we also compute (and cache) the LFD of each cluster using Equation~\ref{eq:methods:lfd-half}.

@@

This complexity analysis ignores the cost of checking the continuation criteria.
While a user could specify criteria of any complexity (our implementation allows for this), for this paper we continue clustering until each cluster is a singleton, i.e.,\,it contains only one point or duplicates of the same point and has a radius of zero.
This criterion costs $\mathcal{O}(1)$ to check per cluster, and so does it not affect the overall complexity of the algorithm as used in this paper.

We also note here that this $\mathcal{O}(n \log n)$ complexity assumes that $\mathcal{T}$ is balanced.
In practice, for real datasets, we expect Algorithm~\ref{alg:methods:partition} to produce anything but a balanced tree; the varying sampling density in different regions of the manifold and the low dimensional ``shape'' of the manifold itself will cause it to be unbalanced.
The only case in which we would expect a balanced tree is if the dataset were uniformly distributed, e.g.,\,in a $d$-dimensional hyper-cube.

@@

% \begin{figure}[H]
%     \vskip -0.2in
%     \centering
%     \includegraphics[scale=0.65]{images/geometry/deltas.pdf}
%     \caption{
%         {\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.
%         ${\color{blue}\delta} = f(q, c)$ is the distance from the query to the cluster center $c$.
%         ${\color{red}\delta^{+}} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
%         ${\color{green}\delta^{-}} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
%     }
%     \label{fig:methods:deltas}
%     \vskip -0.5in
% \end{figure}

\begin{minipage}{.425\textwidth}
    \begin{algorithm}[H]
        \caption{tree-search($C$, $q$, $r$)}
        \label{alg:methods:rnn-search:tree-search}
        \begin{algorithmic}
            \REQUIRE $f$, a distance function
            \REQUIRE $C$, a cluster
            \REQUIRE $q$, a query
            \REQUIRE $r$, a search radius
            \IF{$\delta^+_C \leq r$}
                \STATE \textbf{return} $\{C\}$
            \ELSE
                \STATE $[L, R]$ $\Leftarrow$ \textit{children} of $C$
                \STATE \textbf{return} tree-search($L, q, r$) \\
                \ \ \ \ \ \ \ \ \ \ \ \ \ \ \ $\cup$ tree-search($R, q, r$)
            \ENDIF
        \end{algorithmic}
    \end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{.475\textwidth}
    \begin{algorithm}[H]
        \caption{leaf-search($Q$, $q$, $r$)}
        \label{alg:methods:rnn-search:leaf-search}
        \begin{algorithmic}
            \REQUIRE $f$, a distance function
            \REQUIRE $Q$, a set of clusters
            \REQUIRE $q$, a query
            \REQUIRE $r$, a search radius
            \STATE $H \Leftarrow \emptyset$
            \FOR{$C \in Q$}
                \IF{$\delta^+_C \leq r$}
                    \STATE $H$ $\Leftarrow$ $H \cup C$
                \ELSE
                    \FOR{$p \in C$}
                        \IF{$f(p, q) \leq r$}
                            \STATE $H$ $\Leftarrow$ $H \cup \{p\}$
                        \ENDIF
                    \ENDFOR
                \ENDIF
            \ENDFOR
            \STATE \textbf{return} $H$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}

@@

Given a dataset $\textbf{X}$, a distance function $f$ and the root cluster $\mathcal{R}$ of the tree $\mathcal{T}$ (constructed by Algorithm~\ref{alg:methods:partition}), we can now formally pose the $k$-NN search problem.
Given a user specified integer $k$ and query $q$, $k$-NN search aims to find the $k$ closest points to $q$ in $\textbf{X}$.
In other words, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.

@@

For Repeated $\rho$-NN search (Algorithm~\ref{alg:methods:repeated-rnn}), we begin by performing \textit{tree-search} with a search radius $r$ equal to the radius of the root $\mathcal{R}$ divided by the cardinality of the dataset.
If no clusters are found, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.

@@

\begin{proof} We consider the \textit{tree-search} and \textit{leaf-search} stages of search separately.
Tree-search refers to the process of identifying clusters that overlap with the query ball, or in other words, clusters that might contain one of the $k$ nearest neighbors by doing repeated iterations of the  $\rho$-NN search Algorithm~\ref{alg:methods:rnn-search:tree-search}.
In ~\cite{ishaq2019clustered}, we showed that the complexity of $\rho$-NN search is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{tree-search}}
        \ + \
        \underbrace{
            \overbrace{ \big| B(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ d}^{\textrm{scaling factor}}
        }_{\textrm{leaf-search}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}
where $\hat{r}$ is the \textit{mean} radius of leaf clusters, $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy at that radius, $B(q, \rho)$ is a ball of radius $\rho$ around the query $q$, and $d$ is the LFD around the query at the length scale of $\rho$ and $\rho + 2 \cdot \hat{r}$.

@@

Again using the assumption that since $q$ is from the same distribution as $X$, the LFD at $q$ should not be significantly different from that at a cluster in $Q$, we have that $\hat{d} \approx d$.
By combining the bounds for tree-search and leaf-search, we see that Repeated $\rho$-NN has time complexity as claimed in Equation~\ref{eq:methods:repeated-rnn-complexity}.
\end{proof}

@@

This algorithm (described in Algorithm~\ref{alg:methods:bredth-first-sieve}) performs a breadth-first traversal of the tree $\mathcal{T}$, pruning clusters by using a modified version of the QuickSelect algorithm~\cite{hoare1961algorithm} at each level.

@@

This algorithm (described in Algorithm~\ref{alg:methods:depth-first-sieve}) is quite similar to a depth-first traversal of the tree $\mathcal{T}$, using two priority queues to track clusters and hits, and to prioritize which branch of $\mathcal{T}$ to explore next.

@@

\begin{proof}
Since $q$ is sampled from the same distribution as $X$, the LFD near $q$ should not differ significantly from the LFDs of clusters near $q$ at the scale of the distance from the query to the $k^{th}$ nearest neighbor.

@@

The asymptotic complexity of Breadth-First Sieve is dominated by the QuickSelect algorithm to calculate $\tau$.
Since this method is log-linear in the length of $Q$, and $Q$ contains the clusters from tree-search and the points from leaf-search, we see that the asymptotic complexity is as claimed in Equation~\ref{eq:methods:breadth-first-sieve-complexity}.

For Depth-First Sieve, since we use two priority queues, the asymptotic complexity is as claimed in Equation~\ref{eq:methods:depth-first-sieve-complexity}.
\end{proof}

@@

Based on our asymptotic complexity analyses, we expect CAKES to perform well on datasets with low LFD, and for its performance to scale sub-linearly with the cardinality of the dataset.
To test this hypothesis, we use some datasets from the ANN-benchmarks suite~\cite{aumuller2020ann} and synthetically augment them to generate similar datasets with exponentially larger cardinalities.
We do the same with a large random dataset of uniformly distributed points in a hypercube.
We then compare the performance of CAKES to that of other algorithms on the original datasets and the synthetically augmented datasets.

To elaborate on the augmentation process, we start with an original dataset from the ANN-benchmarks suite.
Let $X$ be the dataset, $d$ be its dimensionality, $\epsilon$ be a user-specified noise level, and $m$ be a user-specified integer multiplier.
For each datum $\mathbf{x} \in X$, we create $m - 1$ new data points within a distance $\epsilon$ of $\mathbf{x}$.
We construct a random vector $\mathbf{r}$ of $d$ dimensions in the hyper-sphere of radius $\epsilon$ centered at the origin.
We then add $\mathbf{r}$ to $\mathbf{x}$ to get a new point $\mathbf{x}'$.
Since $||\mathbf{r}|| \leq \epsilon$, we have that $||\mathbf{x} - \mathbf{x}'|| \leq \epsilon$ (i.e.,\,$\mathbf{x}'$ is within a distance $\epsilon$ of $\mathbf{x}$).
This produces a new dataset $X'$ with $|X'| = m \cdot |X|$.
This augmentation process does not add to the overall topological structure of the dataset, but it does increase its cardinality by a factor of $m$.
This allows us to isolate the effect of cardinality on search performance from that of other factors such as dimensionality, choice of metric, or the topological structure of the dataset.

>>>>

Intuitively, LFD measures the rate of change in the number of points in a ball of radius $r$ around a point $q$ as $r$ increases. When the vast majority of points in the dataset have low ($\ll D$) LFD, we can simply say that the dataset has low LFD.
We stress that this concept differs from the \textit{embedding dimension} of a dataset.
To illustrate the difference, consider the SILVA 18S rRNA dataset that contains genomic sequences with unaligned lengths of up to 3,712 base pairs and aligned length of 50,000 base pairs.
Hence, the \textit{embedding dimension} of this dataset is at least 3,712 and at most 50,000; as used in this paper it is 3,712.
However, physical constraints (namely, biological evolution and biochemistry) constrain the data to a lower-dimensional manifold within this space.
LFD is an approximation of the dimensionality of that lower-dimensional manifold in the ``vicinity'' of a given point.
% Section~\ref{sec:results:lfd-of-datasets} discusses this concept on a variety of datasets, showing how real datasets uphold the manifold hypothesis.
For real-world datasets, we expect the LFD to be locally uniform, i.e.,\,when $r$ is small, but potentially highly variable at larger length scales, i.e.,\,when $r$ is large.

@@

We construct a divisive hierarchical cluster tree $\mathcal{T}$ with CLAM, following CHESS~\cite{ishaq2019clustered} but with improved pole selection, then apply a depth-first reordering (Sec.~\ref{sec:methods:clustering:depth-first-reordering}).

The recursive step is as follows. For any cluster $C$ with $|C|$ points, we draw a random subsample $S\subset C$ of $\sqrt{|C|}$ points, compute all pairwise distances in $S$, and take the geometric median of $S$—the point minimizing the sum of distances to all other points in $S$—as the \emph{center} of $C$. The \emph{radius} of $C$ is the maximum distance from this center to any point in $C$.

Let the \emph{left pole} be the point in $C$ farthest from the center; let the \emph{right pole} be the point farthest from the left pole. We partition $C$ into a \emph{left child} (points closer to the left pole) and a \emph{right child} (points closer to the right pole); ties go to the left child. Starting from the root cluster $\mathcal{R}$ containing the entire dataset, we recurse until each leaf is a single datum or any user-specified stopping criterion is met (e.g., minimum cluster radius, minimum cardinality, or maximum tree depth). During partitioning, we compute and cache the LFD of each cluster using Eq.~\ref{eq:methods:lfd-half}.

@@

@@

% This complexity analysis ignores the cost of checking the continuation criteria.
% While a user could specify criteria of any complexity (our implementation allows for this), for this paper we continue clustering until each cluster is a singleton, i.e.,\,it contains only one point or duplicates of the same point and has a radius of zero.
% This criterion costs $\mathcal{O}(1)$ to check per cluster, and so does it not affect the overall complexity of the algorithm as used in this paper.
We also note here that this $\mathcal{O}(n \log n)$ complexity assumes that $\mathcal{T}$ is balanced.
In practice, for real datasets, we expect the process from Section~\ref{sec:methods:clustering:building-the-tree} to produce anything but a balanced tree; the varying sampling density in different regions of the manifold and the low dimensional ``shape'' of the manifold itself will cause it to be unbalanced.
The only case in which we would expect a balanced tree is if the dataset were uniformly distributed, e.g.,\,in a $d$-dimensional hyper-cube.

@@

\begin{figure}[H]
    \vskip -0.2in
    \centering
    \includegraphics[scale=0.5]{images/geometry/deltas.pdf}
    \caption{
        {\color{blue}$\delta$}, {\color{red}$\delta^{+}$}, and {\color{green}$\delta^{-}$} for a cluster $C$ and a query $q$.
        ${\color{blue}\delta} = f(q, c)$ is the distance from the query to the cluster center $c$.
        ${\color{red}\delta^{+}} = \delta + r$ is the distance from the query to the theoretically farthest point in $C$.
        ${\color{green}\delta^{-}} = \text{max}(0, \delta - r)$ is the distance from the query to the theoretically closest point in $C$.
    }
    \label{fig:methods:deltas}
    \vskip -0.5in
\end{figure}

\begin{minipage}{.425\textwidth}
    \begin{algorithm}[H]
        \caption{tree-search($C$, $q$, $r$)}
        \label{alg:methods:rnn-search:tree-search}
        \begin{algorithmic}
            \REQUIRE $f$, a distance function
            \REQUIRE $C$, a cluster
            \REQUIRE $q$, a query
            \REQUIRE $r$, a search radius
            \IF{$\delta^+_C \leq r$}
                \STATE \textbf{return} $\{C\}$
            \ELSE
                \STATE $[L, R]$ $\Leftarrow$ \textit{children} of $C$
                \STATE \textbf{return} tree-search($L, q, r$) \\
                 $\cup$ tree-search($R, q, r$)
            \ENDIF
        \end{algorithmic}
    \end{algorithm}
\end{minipage}
\hfill
\begin{minipage}{.475\textwidth}
    \begin{algorithm}[H]
        \caption{leaf-search($Q$, $q$, $r$)}
        \label{alg:methods:rnn-search:leaf-search}
        \begin{algorithmic}
            \REQUIRE $f$, a distance function
            \REQUIRE $Q$, a set of clusters
            \REQUIRE $q$, a query
            \REQUIRE $r$, a search radius
            \STATE $H \Leftarrow \emptyset$
            \FOR{$C \in Q$}
                \IF{$\delta^+_C \leq r$}
                    \STATE $H$ $\Leftarrow$ $H \cup C$
                \ELSE
                    \FOR{$p \in C$}
                        \IF{$f(p, q) \leq r$}
                            \STATE $H$ $\Leftarrow$ $H \cup \{p\}$
                        \ENDIF
                    \ENDFOR
                \ENDIF
            \ENDFOR
            \STATE \textbf{return} $H$
        \end{algorithmic}
    \end{algorithm}
\end{minipage}

@@

Let $\textbf{X}$ be a dataset, $f$ a distance function, and $\mathcal{R}$ the root cluster of a tree $\mathcal{T}$ constructed by the process in Section~\ref{sec:methods:clustering:building-the-tree}. Given a user-specified integer $k$ and query $q$, $k$-NN search aims to find the $k$ closest points to $q$ in $\textbf{X}$.
In other words, $k$-NN search aims to find the set $H$ such that $|H| = k$ and $H = B_X(q, \rho_k)$ where $\rho_k = \max \left\{ f(q, p) \ \forall \ p \in H \right\}$ is the distance from $q$ to the $k^{th}$ nearest neighbor in $\textbf{X}$.

@@

For Repeated $\rho$-NN search, we begin by performing \textit{tree-search} with a search radius $r$ equal to the radius of the root $\mathcal{R}$ divided by the cardinality of the dataset.
If no clusters are found, then we double $r$ and perform tree-search again, repeating until we find at least one cluster.

@@

We consider the \textit{tree-search} and \textit{leaf-search} stages of search separately.
Tree-search refers to the process of identifying clusters that overlap with the query ball, or in other words, clusters that might contain one of the $k$ nearest neighbors by doing repeated iterations of the  $\rho$-NN search Algorithm~\ref{alg:methods:rnn-search:tree-search}.
In ~\cite{ishaq2019clustered}, we showed that the complexity of $\rho$-NN search is
\begin{gather}
    \mathcal{O}
    \Bigg(
        \underbrace{
            \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
        }_{\textrm{tree-search}}
        \ + \
        \underbrace{
            \overbrace{ \big| B(q, \rho) \big|}^{\textrm{output size}}
            \overbrace{ \left( \frac{\rho + 2 \cdot \hat{r}}{ \rho} \right) ^ d}^{\textrm{scaling factor}}
        }_{\textrm{leaf-search}}
    \Bigg)
    \label{eq:methods:rnn-search-complexity}
\end{gather}
where $\hat{r}$ is the \textit{mean} radius of leaf clusters, $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy at that radius, $B(q, \rho)$ is a ball of radius $\rho$ around the query $q$, and $d$ is the LFD around the query at the length scale of $\rho$ and $\rho + 2 \cdot \hat{r}$.

@@

Again using the assumption that since $q$ is from the same distribution as $X$, the LFD at $q$ should not be significantly different from that at a cluster in $Q$, we have that $\hat{d} \approx d$.
By combining the bounds for tree-search and leaf-search, we see that time complexity of performing Repeated $\rho$-NN search on $X$ with query $q$ is \begin{gather}
            \mathcal{O}
            \Bigg(
                \underbrace{
                    \log~\overbrace{\mathcal{N}_{\hat{r}}(X)}^{\textrm{metric entropy}}
                }_{\textrm{tree-search}}
                \ + \
                \underbrace{
                    \overbrace{k}^{\textrm{output size}} \cdot
                    \overbrace{\bigg( 1 + 2 \cdot \Big( \frac{\hat{|C|}}{k} \Big) ^ {d^{-1}} \bigg)^d}^{\textrm{scaling factor}}
                }_{\textrm{leaf-search}}
            \Bigg)
            \label{eq:methods:repeated-rnn-complexity}
        \end{gather}
        where $\mathcal{N}_{\hat{r}}(X)$ is the metric entropy of the dataset, $d$ is the LFD of the dataset, and $k$ is the number of nearest neighbors.

@@

This algorithm performs a breadth-first traversal of the tree $\mathcal{T}$, pruning clusters by using a modified version of the QuickSelect algorithm~\cite{hoare1961algorithm} at each level.

@@

This algorithm is quite similar to a depth-first traversal of the tree $\mathcal{T}$, using two priority queues to track clusters and hits, and to prioritize which branch of $\mathcal{T}$ to explore next.

@@

Since $q$ is sampled from the same distribution as $X$, the LFD near $q$ should not differ significantly from the LFDs of clusters near $q$ at the scale of the distance from the query to the $k^{th}$ nearest neighbor.

@@

The asymptotic complexity of Breadth-First Sieve is dominated by the QuickSelect algorithm to calculate $\tau$.
Since this method is log-linear in the length of $Q$, and $Q$ contains the clusters from tree-search and the points from leaf-search, we see that the asymptotic complexity of Breadth-First Sieve is $\mathcal{O} \Big( (T + L ) \log (T + L ) \Big) \label{eq:methods:breadth-first-sieve-complexity}$
and the the time complexity of Depth-First Sieve search is $ \mathcal{O} \Big( T \log T + L \log k \Big).
            \label{eq:methods:depth-first-sieve-complexity} $

@@

Based on our asymptotic analyses, we expect CAKES to perform well on datasets with low LFD and to scale sublinearly with dataset cardinality. To test this, we use datasets from the ANN-Benchmarks suite~\cite{aumuller2020ann} and synthetically augment them to produce similar datasets with exponentially larger cardinalities. We perform the same procedure on a large dataset of uniformly distributed points in a hypercube, and compare CAKES with other algorithms on both the original and augmented datasets.

Our augmentation procedure is as follows. Let $X$ be a dataset of dimension $d$, $\epsilon$ a user-specified noise level, and $m$ an integer multiplier. For each $\mathbf{x}\in X$, generate $m-1$ new points within distance $\epsilon$ of $\mathbf{x}$ by sampling a random vector $\mathbf{r}\in\mathbb{R}^d$ from the $\epsilon$-ball (hypersphere of radius $\epsilon$ centered at the origin) and setting $\mathbf{x}'=\mathbf{x}+\mathbf{r}$. Since $\lVert\mathbf{r}\rVert\le\epsilon$, we have $\lVert\mathbf{x}-\mathbf{x}'\rVert\le\epsilon$. The result is $X'$ with $\lvert X'\rvert=m\,\lvert X\rvert$. This increases cardinality by a factor of $m$ without altering the dataset’s overall topological structure, isolating the effect of cardinality from dimensionality, metric choice, and topology.

### 2025-08-26 sections/3-datasets.tex
version where we don't discuss lfd
tightening discussion

In addition to benchmarks on datasets from the ANN-Benchmarks suite, we also benchmarked on synthetic augmentations of these real datasets, using the process described in \ref{sec:methods:synthetic-data}.
In particular, we use a noise tolerance $\epsilon = 0.01$ and explore the scaling behavior as the cardinality multiplier (referred to as ``Mult.'' in Tables~\ref{tab:results:qps-and-recall-fmn},~\ref{tab:results:qps-and-recall-glove},~\ref{tab:results:qps-and-recall-sift} and~\ref{tab:results:qps-and-recall-random}) increases.
We also benchmarked on purely randomly-generated datasets of various cardinalities.
For this, we used a base cardinality of 1,000,000 and a dimensionality of 128 to match the Sift dataset; hereafter, we refer to the random dataset with cardinality 1,000,000 and dimensionality 128 as ``Random.''
This benchmark allows us to isolate the effect of a manifold structure (which we expect to be absent in a purely random dataset) on the performance of the CAKES' algorithms.

In order to calculate recall on the augmented datasets, we perform linear search in Rust and save the results to disk.
We verified that linear search on the original dataset produced neighbors with perfect recall compared to those provided by the ANN-Benchmarks suite.
We then use the results of linear search on the augmented datasets as ground truth for calculating recall of CAKES's algorithms in Rust and read them in Python for calculating recall of HNSW, ANNOY, and FAISS-IVF.

@@

To demonstrate CAKES with a more exotic distance function, we also benchmarked on the SILVA 18S ribosomal RNA dataset~\cite{10.1093/nar/gks1219}.
This dataset contains ribosomal RNA sequences of 2,224,640 genomes, with the longest sequence having 3,712 letters.
We held out a set of 1,000 random sequences from the dataset to use as queries for benchmarking.
We use Levenshtein~\cite{levenshtein1966binary} distance on the unaligned sequences to build the tree and to perform $k$-NN search.
We note that the sequences in this dataset were provided in a multiple sequence alignment with a width of 50,000 characters.
We could have used Hamming distance on the aligned sequences, but we chose to use Levenshtein distance on the unaligned sequences to help demonstrate the flexibility of CAKES in handling exotic distance functions.
Under Hamming distance, we could consider this dataset to have an embedding dimension of 50,000, but for Levenshtein distance, we state the dimensionality as the length of the longest sequence, i.e.\, 3,712.

@@

As another example of an exotic distance function, we benchmarked CAKES on the Radio-ML dataset~\cite{oshea2018radioml} under Dynamic Time Warping~\cite{muller2007dynamic} as the distance function.
This dataset contains samples of synthetically generated signal captures of different modulation modes over a range of signal-to-noise ratio (SNR) levels.
Specifically, it comprises 24 modulation modes at 26 different SNRs ranging from -20 dB to 30 dB, with 4,096 samples at each combination of modulation mode and SNR level.
Thus, it contains $24 \cdot 26 \cdot 4096 = 2,555,504$ samples in total.
Each sample is a 1,024-dimensional complex-valued vector, representing a signal capture (a time-series of complex-valued numbers).
We used a subset of this dataset, containing $97,304$ samples at 10dB SNR and used another $1,000$ samples at the same SNR as a hold-out set of queries.

>>>>

In addition to ANN-Benchmarks datasets, we evaluate synthetic augmentations generated as in Sec.~\ref{sec:methods:synthetic-data}, using a noise tolerance $\epsilon=0.01$, and study scaling as the cardinality multiplier (“Mult.” in Tables~\ref{tab:results:qps-and-recall-fmn}, \ref{tab:results:qps-and-recall-glove}, \ref{tab:results:qps-and-recall-sift}, \ref{tab:results:qps-and-recall-random}) increases. We also benchmark purely random datasets at various cardinalities; to match Sift we use a random dataset with base cardinality of $1{,}000{,}000$ and dimensionality $128$, and refer to this dataset as “Random.” This benchmark isolates the effect of manifold structure—expected to be absent in purely random data—on CAKES algorithms’ performance.

To compute recall on the augmented datasets, we run linear search in Rust and write the results to disk. We verified on the ANN-Benchmark datasets that this linear search attains perfect recall using the test data provided. We then use the linear-search results on the augmentations as ground truth to evaluate CAKES in Rust and to compute recall for HNSW, ANNOY, and FAISS-IVF in Python.

@@

To demonstrate CAKES with an exotic distance, we benchmark the SILVA 18S ribosomal RNA dataset~\cite{10.1093/nar/gks1219}, which contains rRNA sequences from 2,224,640 genomes; the longest sequence has 3,712 letters. We hold out 1,000 random sequences as queries. We build the tree and perform $k$-NN using Levenshtein distance~\cite{levenshtein1966binary} on unaligned sequences. Although the dataset is provided as a multiple sequence alignment of width 50{,}000, we intentionally use Levenshtein rather than Hamming on the aligned strings to demonstrate CAKES’s flexibility. Under Hamming, the embedding dimension would be 50{,}000; under Levenshtein, we report the dimensionality as the length of the longest sequence (3{,}712).

@@

We also benchmark CAKES on the Radio-ML dataset~\cite{oshea2018radioml} using Dynamic Time Warping~\cite{muller2007dynamic} as the distance. The dataset spans 24 modulation modes across 26 SNR levels from $-20$ to $30$\,dB, with 4{,}096 samples per combination, totaling $24\!\cdot\!26\!\cdot\!4096=2{,}555{,}504$ samples. Each sample is a 1{,}024-dimensional complex-valued vector (a time series). For experiments, we use the 10dB SNR subset with 97{,}304 samples and a 1{,}000-sample hold-out query set at the same SNR.

% \subsection{Local Fractal Dimension}
% \label{sec:dayasets:lfd-of-datasets}

% Since the CAKES algorithms were designed to scale in time with the LFD of the dataset, we examine the LFD of each dataset we used for benchmarks.
% Figure~\ref{fig:results:lfd-plots} illustrates the trends in LFD for Fashion-MNIST, Glove-25, Sift, Random, Silva 18S, and Radio-ML.
% In this section, when we discuss trends in LFD, unless otherwise noted, we are referring to the 95$^{th}$ percentile of LFD.

% For the Fashion-MNIST dataset we observe in Figure~\ref{fig:results:fashion-mnist-lfd} that until approximately depth 5, Fashion-MNIST's LFD is low (i.e., less than 4).
% It then starts increasing, reaching a peak of about 6 near depth 20, before decreasing to 1 at the maximum depth.

% Relative to Fashion-MNIST, Glove-25 has low LFD, as shown in Figure~\ref{fig:results:glove-25-lfd}.
% In particular, Glove-25's LFD is less than 3 for all depths.

% Figure~\ref{fig:results:sift-lfd} shows the LFD by depth for Sift, which has higher LFD relative to Fashion-MNIST and Glove-25.
% It increases sharply to a peak of 9 around a depth of 10, after which it decreases smoothly until reaching the deepest leaves in the tree.

% We contrast the low LFD of Sift with that of the Random dataset which has the same cardinality and dimension. As shown in Figure~\ref{fig:results:random-lfd}, the character of this dataset is very different from the others.
% The LFD starts at 20 at depth 0 and all percentile lines decrease linearly with depth until reaching the leaves of the tree.
% The spread in LFD starts very small for the first few clusters and increases as depth increases.

% The Silva dataset, shown in Figure~\ref{fig:results:silva-lfd}, exhibits consistently low LFD.
% In particular, LFD is less than 3 for all depths, hovering near 1 for clusters at depth 40 and deeper.

% For the Radio-ML dataset, the LFD values show three distinct peaks around an LFD of 12 at or near depths of 8, 25 and 50.
% Each peak is followed by a linear decrease until encountering a sharp spike for the next peak.
% Within each of the tree portions, this dataset has a character similar to that of the Random dataset.
% This suggests that the dataset obeys the manifold hypothesis at some scales, but that it is not ``scale free,'' as the LFD varies significantly by depth.
% This is likely the result of a piecewise uniform sampling strategy used to generate the different modulation modes present in the dataset.
